Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-02-15 text/paper2/ms.tex
references and intro text

\begin{abstract}
Context: Connecting cosmological simulations to observed galaxies is difficult because baryonic processes not well understood/modeled but also because they are gravitationally dominated by unobservable dark matter. 
Aims: Develop a method for inferring higher order properties of the Galactic potential using debris from disrupted satellites.
Methods: 
Results: 
Conclusions:

@@

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the major observable components of a galaxy are sensitive to the inner portion of its host halo's mass distribution. For example, the rotation curves of disk galaxies are excellent probes of the inner mass, especially since matter in disks can be assumed to move on near circular orbits. The inner potential is likely well-described by a closely-integrable potential, thus allowing [...] Conversely, measuring the dark matter distribution at large radii is complicated by the low density of visible tracers and [...something about assumptions about integrability and smoothness]. Around external galaxies, the extended mass distribution has been probed using a variety of approaches \citep[see][for a a complete and detailed review]{courteau13}. For example, the kinematics of tracer populations such as globular clusters or planetary nebulae can be used to derive mass estimates under the assumptions that these satellite systems are relaxed and well-mixed \citep[early investigations include][]{mendez01,cote03}. Simple, parameterized models of both the mass and orbit distribution have been simultaneously constrained using such data \citep[e.g.][]{napolitano11,deason12c}. 
%\citep{romanowsky03,douglas07,romanowsky09,napolitano11,lee11}
Alternatively, the statistical properties of gravitationally lensed background sources around a galaxy can be used to constrain the \emph{projected} shape, orientation, and radial profile of mass \citep[as done by the Lens Structure and Dynamics Survey described in][]{koopmans02}. Of course, lensing reconstructions can only be performed for galaxies which closely intersect our line of sight to background sources, but the advent of large photometric catalogues has allowed automatic searches for such chance alignments and significant increases in the number of objects studied in this way \citep[e.g. the Sloan Lens ACS Survey, see][]{bolton06}.

Within the Milky Way our unique vantage point allows us a three-dimensional view of the structure in own dark matter halo. Our proximity allows us to use individual stars as kinematic tracers and hence build much larger samples that probe deeper into the halo than the globular cluster and planetary nebula studies of external galaxies. For example, \cite{deason12a} used halo BHB stars selected from the Sloan Digital Sky Survey \cite[SDSS;][]{york00} as a random tracer population to measure the mass and slope of a simple, power-law fit to the Galactic potential. Such studies assume that the tracer orbits are randomly sampled from a smooth distribution function and phase mixed, however the existence of large-scale substructure in stellar streams and clouds is known to bias mass and velocity inferences that ignore this fact. Another approach is to exploit the non-random nature of the stellar populations in the halo. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might assume that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.,][]{}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g., along a stream), which would give us a direct measure of differences in a potential, $\Phi$. 
%In practice, simple methods such as orbit fitting lead to systematic biases in inferred properties of the underlying potential because streams do not strictly delineate orbits [cite eyre binney work].

@@

[Here we need to talk about integrable potentials and shortcomings of the methods above?] Methods that rely on action-angle coordinates, [...]. Any generative model that relies on matching densities --- whether in phase-space, action-space, or observed coordinates --- must also include details about the progenitor and a model for the background of stars. [...] 
- Then a short note on worries\\
\indent\indent- for generative models\\
\indent\indent- for any methods based on action/angle \\
\indent\indent- GD-1 was fit as if it highlighted an orbit.  That is known to be wrong.\\
\indent\indent- Densities are hard to match: need a background model, way to go from luminosity to density, needs assumptions about the mass function and structure of the satellite\\
\indent\indent- Triaxiality / radially-dependent shapes and substructure present problems for integral-based methods\\
- Then how rewinder gets round these????\\
\indent\indent- Reference Deg \& Widrow to say they illustrate the idea of the power of multiple streams to back up the bit where you say that's the long-term aim

[Actions, angles, integrable, not, etc.  Can integrate in any
  potential, even a time-varying one. We use this to
  ``run the clock back''.  Why is an objective based on past collisions not the
  same as assuming that the stars are on the same orbit?]

@@

\section{[Scalings?]}

In \paperone, we define a set of coordinates relative to the instantaneous phase-space position of the progenitor, $(\bs{x}_{p},\bs{v}_{p})$, normalized by the instantaneous tidal radius and escape velocity. Here we (1) instead scale the velocity coordinates by the velocity dispersion of the satellite instead of the escape velocity, (2) reduce the dimensionality by considering only the magnitude of these vectors, and (3) take the logarithm of the magnitude,
\begin{equation}
  R = \ln\left(\frac{\left\vert\bs{x} - \bs{x}_{p}\right\vert}{R_{\rm tide}}\right)\,\,\,\,,\,\,\,\,
  V = \ln\left(\frac{\left\vert\bs{v} - \bs{v}_{p}\right\vert}{\sigma_v}\right).
\end{equation}

>>>>

\begin{abstract}
Context: Connecting cosmological simulations to observed galaxies is difficult because baryonic processes not well understood/modeled but also because they are gravitationally dominated by unobservable dark matter. One powerful measure is the shape of the potential -- triaxial, density profile, etc. Tidal streams are excellent tracers of mass.

Aims: Present a method for inferring higher order properties of the Galactic potential using debris from disrupted satellites.

Methods:

Results: Can simultaneously constrain properties of progenitor, missing dimensions, and potential with a simple model for disrupted debris in phase-space. The method works well with small numbers of stars and with finite or large observational uncertainties or missing dimensions and generalizes naturally to multiple independent debris structures.

Conclusions:

@@

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the major observable components of a galaxy are sensitive to the inner portion of its host halo's mass distribution. For example, the rotation curves of disk galaxies are excellent probes of the inner mass, especially since matter in disks can be assumed to move on near circular orbits. Conversely, measuring the dark matter distribution at large radii is complicated by the low density of visible tracers, observational difficulties, and [unclear what assumptions can be made about the integrability and smoothness of the mass distribution]. Around external galaxies, the extended mass distribution has been probed using a variety of approaches \citep[see][for a a complete and detailed review]{courteau13}. For example, the kinematics of tracer populations such as globular clusters or planetary nebulae can be used to derive mass estimates under the assumptions that these satellite systems are relaxed and well-mixed \citep[early investigations include][]{mendez01,cote03}. Simple, parameterized models of both the mass and orbit distribution have been simultaneously constrained using such data \citep[e.g.][]{napolitano11,deason12c}. 
%\citep{romanowsky03,douglas07,romanowsky09,napolitano11,lee11}
Alternatively, the statistical properties of gravitationally lensed background sources around a galaxy can be used to constrain the \emph{projected} shape, orientation, and radial profile of mass \citep[as done by the Lens Structure and Dynamics Survey described in][]{koopmans02}. Of course, lensing reconstructions can only be performed for galaxies which closely intersect our line of sight to background sources, but the advent of large photometric catalogues has allowed automatic searches for such chance alignments and significant increases in the number of objects studied in this way \citep[e.g. the Sloan Lens ACS Survey, see][]{bolton06}.

Within the Milky Way our unique vantage point allows us a three-dimensional view of the structure in own dark matter halo. Our proximity allows us to use individual stars as kinematic tracers and hence build much larger samples that probe deeper into the halo than the globular cluster and planetary nebula studies of external galaxies. For example, \cite{deason12a} used halo BHB stars selected from the Sloan Digital Sky Survey \cite[SDSS;][]{york00} as a random tracer population to measure the mass and slope of a simple, power-law fit to the Galactic potential. Such studies assume that the tracer orbits are randomly sampled from a smooth distribution function and phase mixed, however the existence of large-scale substructure in stellar streams and clouds is known to bias mass and velocity inferences that ignore this fact. Another approach is to exploit the non-random nature of the stellar populations in the halo. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might assume that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.,][]{eyre2011}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g., along a stream), which would give us a direct measure of differences in a potential, $\Phi$. 
%In practice, simple methods such as orbit fitting lead to systematic biases in inferred properties of the underlying potential because streams do not strictly delineate orbits [cite eyre binney work].

@@

The potential of the inner galaxy is likely well-described by a (set of) closely-integrable potentials, thus the dynamics of collisionless systems is plausibly captured by a transformation to some integral-of-motion-space, e.g., action-angle coordinates. However there is uncertainty in the exact shape, radial profile, and graininess of the outer halo. To test the extent to which the outer halo is well-fit by integrable potentials, we cannot use methods that \emph{assume} integrability. In this article, we present a fully probabilistic method for measuring properties of a potential that only relies on numerical orbit integration in ordinary phase-space. The model uses individual stars as constraints on the potential, and can therefore handle very small samples of stars. Generative models that instead rely on matching densities (in phase-space, action-space, or observed coordinates) must also include fine details about the progenitor (e.g., mass function or structure) and a model for the background of stars. Our method also naturally extends to using multiple debris structures \citep[see][who illustrate the power of using multiple streams to simultaneously constrain the potential]{deg14}.

@@

\section{Scalings}

In \paperone, we define a set of coordinates relative to the instantaneous phase-space position of the progenitor, $(\bs{x}_{p},\bs{v}_{p})$, normalized by the instantaneous tidal radius and escape velocity. Here we (1) scale the velocity coordinates by the velocity dispersion of the satellite instead of the escape velocity, (2) reduce the dimensionality by considering only the magnitude of these vectors, and (3) take the logarithm of the magnitude,
\begin{equation}
  R = \ln\left(\frac{\left\vert\bs{x} - \bs{x}_{p}\right\vert}{R_{\rm tide}}\right)\,\,\,\,,\,\,\,\,
  V = \ln\left(\frac{\left\vert\bs{v} - \bs{v}_{p}\right\vert}{\sigma_v}\right).
\end{equation}
